---
title: "Customer Segmentation Analysis (RFM + Clustering)"
author: "Javier Kangditya"
output: html_document
---

# Introduction

## Project Overview
This project analyzes customer transaction behavior and segments customers using the RFM (Recency, Frequency, Monetary) framework combined with clustering techniques.

The objective is to uncover meaningful customer groups to support data-driven marketing and retention strategies.

Unlike predictive modeling tasks, this project focuses on unsupervised learning to identify natural patterns in customer behavior.

---

## Dataset Description
The dataset contains retail transaction records with the following variables:

- InvoiceNo  
- StockCode  
- Description  
- Quantity  
- InvoiceDate  
- UnitPrice  
- CustomerID  
- Country  

This dataset represents real-world transactional behavior suitable for customer segmentation analysis.

---

## Project Goals
- Clean and prepare transactional data  
- Construct RFM features per customer  
- Perform exploratory data analysis (EDA)  
- Apply clustering to segment customers  
- Interpret customer segments from a business perspective  

---

# 1. Data Cleaning
```{r data_cleaning, message=FALSE, warning=FALSE}
# Load necessary libraries
library(tidyverse)
library(lubridate)

# Load dataset
data <- read.csv("Data/Raw/data.csv")

# Preview dataset
glimpse(data)
head(data)

# Placeholder: handle missing values, duplicates, and correct data types

# Remove duplicates
data <- data %>% distinct()

# Drop rows with missing CustomerID
data <- data %>% drop_na(CustomerID)

# Remove cancelled / invalid transactions
data <- data %>%
  filter(Quantity > 0, UnitPrice > 0)

# Create TotalPrice feature
data <- data %>%
  mutate(TotalPrice = Quantity * UnitPrice)

# Convert InvoiceDate to datetime
data$InvoiceDate <- as.POSIXct(data$InvoiceDate, format = "%m/%d/%Y %H:%M")

# Check cleaned data
glimpse(data)

library(openxlsx)

# Save processed data
write.csv(data, 
          "Data/Processed/retail_data_clean.csv", 
          row.names = FALSE)
```

---

# 2. RFM Feature Engineering
```{r rfm_feature_engineering, message=FALSE, warning=FALSE}

# Load library
library(dplyr)
library(lubridate)

# -----------------------------
# 1. Set snapshot date
# -----------------------------
snapshot_date <- max(data$InvoiceDate) + days(1)

# -----------------------------
# 2. Build RFM table
# -----------------------------
rfm <- data %>%
  group_by(CustomerID) %>%
  summarise(
    Recency = as.numeric(snapshot_date - max(InvoiceDate)),
    Frequency = n_distinct(InvoiceNo),
    Monetary = sum(TotalPrice)
  )

# -----------------------------
# 3. Inspect RFM result
# -----------------------------
glimpse(rfm)
head(rfm)

# Check summary statistics
summary(rfm)

# Save RFM table
write.csv(rfm,
          "Data/Processed/rfm_table.csv",
          row.names = FALSE)
```        

---

# 3. Exploratory Data Analysis (EDA)
```{r eda, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)

# -----------------------------
# 1. Summary statistics
# -----------------------------
summary(rfm)

# -----------------------------
# 2. Recency Distribution
# -----------------------------
p_recency <- ggplot(rfm, aes(x = Recency)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(
    title = "Distribution of Recency",
    x = "Recency (Days)",
    y = "Number of Customers"
  )

p_recency

ggsave("Outputs/Figures/recency_distribution.png",
       plot = p_recency,
       width = 6, height = 4, dpi = 300)

# -----------------------------
# 3. Frequency Distribution
# -----------------------------
p_frequency <- ggplot(rfm, aes(x = Frequency)) +
  geom_histogram(bins = 100, fill = "darkgreen", color = "white") +
  theme_minimal() +
  labs(
    title = "Distribution of Frequency",
    x = "Frequency (Number of Purchases)",
    y = "Number of Customers"
  )

p_frequency

ggsave("Outputs/Figures/frequency_distribution.png",
       plot = p_frequency,
       width = 6, height = 4, dpi = 300)

# -----------------------------
# 4. Monetary Distribution
# -----------------------------
p_monetary <- ggplot(rfm, aes(x = Monetary)) +
  geom_histogram(bins = 100, fill = "orange", color = "white") +
  theme_minimal() +
  labs(
    title = "Distribution of Monetary Value",
    x = "Monetary Value",
    y = "Number of Customers"
  )

p_monetary

ggsave("Outputs/Figures/monetary_distribution.png",
       plot = p_monetary,
       width = 6, height = 4, dpi = 300)

# -----------------------------
# 5. Boxplots
# -----------------------------
p_box_r <- ggplot(rfm, aes(y = Recency)) +
  geom_boxplot(fill = "steelblue") +
  theme_minimal() +
  labs(title = "Outliers in Recency")

p_box_r

ggsave("Outputs/Figures/recency_boxplot.png",
       plot = p_box_r,
       width = 6, height = 4, dpi = 300)

p_box_f <- ggplot(rfm, aes(y = Frequency)) +
  geom_boxplot(fill = "darkgreen") +
  theme_minimal() +
  labs(title = "Outliers in Frequency")

p_box_f

ggsave("Outputs/Figures/frequency_boxplot.png",
       plot = p_box_f,
       width = 6, height = 4, dpi = 300)

p_box_m <- ggplot(rfm, aes(y = Monetary)) +
  geom_boxplot(fill = "orange") +
  theme_minimal() +
  labs(title = "Outliers in Monetary Value")

p_box_m

ggsave("Outputs/Figures/monetary_boxplot.png",
       plot = p_box_m,
       width = 6, height = 4, dpi = 300)

# -----------------------------
# 6. Correlation
# -----------------------------
cor_matrix <- cor(rfm[, c("Recency", "Frequency", "Monetary")])
cor_matrix
```

---

# 4. Data Scaling for Clustering
```{r data_scaling, message=FALSE, warning=FALSE}
library(dplyr)

# Select RFM features only
rfm_features <- rfm %>%
  select(Recency, Frequency, Monetary)

# Standardize the data
rfm_scaled <- scale(rfm_features)

# Convert to data frame 
rfm_scaled <- as.data.frame(rfm_scaled)

# Check result
summary(rfm_scaled)

write.csv(rfm_scaled,
          "Data/Processed/rfm_scaled.csv",
          row.names = FALSE)
```

---

# 5. Elbow Method
```{r elbow_method, message=FALSE, warning=FALSE}

set.seed(123)

wss <- vector()

# Try k from 1 to 10
for (k in 1:10) {
  kmeans_model <- kmeans(rfm_scaled, centers = k, nstart = 25)
  wss[k] <- kmeans_model$tot.withinss
}

# Convert to data frame
elbow_data <- data.frame(
  k = 1:10,
  wss = wss
)

# Plot Elbow Curve
library(ggplot2)

ggplot(elbow_data, aes(x = k, y = wss)) +
  geom_line(color = "steelblue") +
  geom_point(color = "red") +
  theme_minimal() +
  labs(
    title = "Elbow Method for Optimal K",
    x = "Number of Clusters (k)",
    y = "Within-Cluster Sum of Squares (WSS)"
  )

# Save Elbow Method Plot
ggsave("Outputs/Figures/elbow_method.png",
       width = 6, height = 4, dpi = 300)
```

---

# 6. K-Means Clustering
```{r kmc, message=FALSE, warning=FALSE}
set.seed(123)

k_final <- 4

kmeans_model <- kmeans(rfm_scaled, centers = k_final, nstart = 25)

# Add cluster label to original RFM data
rfm$Cluster <- as.factor(kmeans_model$cluster)

# Check result
head(rfm)

# Save RFM with Cluster Labels
write.csv(rfm,
          "Data/Processed/rfm_with_clusters.csv",
          row.names = FALSE)
saveRDS(kmeans_model,
        "Outputs/Models/kmeans_model.rds")
```

---

# 7. Cluster Interpretation
```{r cluster_interpretation, message=FALSE, warning=FALSE}
cluster_summary <- rfm %>%
  group_by(Cluster) %>%
  summarise(
    avg_recency = mean(Recency),
    avg_frequency = mean(Frequency),
    avg_monetary = mean(Monetary),
    count = n()
  ) %>%
  arrange(avg_recency)

cluster_summary

write.csv(cluster_summary,
          "Data/Processed/cluster_summary.csv",
          row.names = FALSE)
```

---

# 8. Cluster Visualization
```{r cluster_visualization, message=FALSE, warning=FALSE}
library(ggplot2)

cluster_plot <- ggplot(rfm, aes(x = Frequency, y = Monetary, color = Cluster)) +
  geom_point(alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "Customer Segments based on K-Means Clustering",
    x = "Frequency (Purchase Count)",
    y = "Monetary Value"
  )

cluster_plot

ggsave("Outputs/Figures/kmeans_cluster_fm.png",
       plot = cluster_plot,
       width = 6, height = 4, dpi = 300)
```

---

# Conclusion & Insights

## 1. Summary of Findings

- **Exploratory Data Analysis (EDA):**  
  - Customer behavior shows strong skewness in `Frequency` and `Monetary`, indicating that a small group of customers contributes most of the revenue.  
  - `Recency` reveals that a significant portion of customers have not purchased recently, indicating potential churn risk.  
  - Outliers exist in `Monetary`, representing high-value customers (VIP segment).

- **RFM Feature Engineering:**  
  - RFM variables (`Recency`, `Frequency`, `Monetary`) were successfully constructed from transactional data.  
  - These features provide a behavioral view of customers based on purchase timing, frequency, and spending value.

- **Clustering Results (K-Means):**  
  - Customers were segmented into multiple clusters using K-Means clustering.  
  - Each cluster shows different behavioral patterns in terms of recency, frequency, and monetary value.  
  - A small group of customers represents **high-value VIP customers**, while the largest group consists of **low-value or occasional customers**.

## 2. Practical Implications

- A small segment of customers contributes disproportionately to revenue, showing a clear **Pareto (80/20) effect**.  
- **VIP customers** should be prioritized for retention strategies such as loyalty programs and exclusive offers.  
- **Churn-risk customers** (high Recency) should be targeted with reactivation campaigns.  
- **Occasional customers** represent growth potential through targeted promotions and engagement strategies.  
- RFM segmentation provides a simple and effective framework for customer-centric marketing decisions.

## 3. Recommendations

- Focus retention strategies on **VIP customers** to maintain high revenue contribution.  
- Implement **win-back campaigns** for inactive customers with high Recency values.  
- Use targeted marketing campaigns to increase purchase frequency among occasional customers.  
- Regularly update RFM segmentation to reflect changing customer behavior.  
- Integrate this segmentation approach into CRM systems for data-driven marketing strategies.